Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-29 parallel-tractability-J-v9/parallel-tractability-J.tex
abstract shortened

\address[add2]{
    Institution of Artificial Intelligence, University of Ulm
}

\begin{abstract}
Materialization is an important reasoning service for applications built
on the ontology languages, such as OWL. With a
rapid growth of semantic data, it is challenging to perform
materialization on large-scale ontologies efficiently.
Current work proposes parallel algorithms and employs computing
platforms of high-performance to make materialization sufficiently efficient and scalable in practice.
However, there do exist ontologies, for which
materialization cannot always be improved by parallelism (we use experiments to show this in Section~\ref{sec:evaluation}).
On the other hand, some well-known large-scale ontologies, such as YAGO, have been shown to have
good performance for parallel reasoning, but they are expressed in ontology languages
that are not parallelly tractable in theory, i.e., the efficiency of reasoning may not be improved
on a parallel implementation.
This motivates us to study the problem of \emph{parallel tractability} of ontology
materialization from a theoretical perspective.
That is, we aim to identify the ontologies for which materialization is
parallelly tractable, i.e., in NC complexity.
The results of the parallel tractability can be further used to optimize
materialization algorithms and to guide ontology engineers in creating parallelly tractable ontologies.

In this work, we focus on datalog rewritable ontology languages.
We propose some algorithms, called NC algorithms, to identify several classes of datalog rewritable
ontologies (called \emph{parallelly tractable classes})
such that materialization over them is parallelly tractable.
We then investigate the parallel tractability of materialization over
two datalog rewritable ontology languages DL-Lite and DHL (\emph{Description Horn Logic}).
We show that materialization over any ontology expressed in DL-Lite$_{core}$ or DL-Lite$_\mathcal{R}$ is parallelly tractable.
For DHL, there exist ontologies where materialization can hardly be parallelized.
We propose to restrict the usage of DHL such that the restricted ontologies are parallelly tractable.
We further extend the results to an extension of DHL that also allows \emph{complex role inclusion axioms}.
To verify the practical usability of the above results, we analyze real-world datasets and show
that many ontologies expressed in DHL or its extension belong to
the parallelly tractable classes. Based on the results of parallel tractability,
we further implement an optimized parallel algorithm
and compare it to the state-of-the-art reasoner RDFox. The experimental results show that the
optimization used in our implementation results in a better
performance on parallelly tractable ontologies compared with RDFox.
\end{abstract}

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep NC complexity
\end{keyword}

@@

There has been work proposing parallel reasoning algorithms and employing parallel computing platforms
for the ontology language RDFS and its extended
fragments \cite{MotikNPHO14,PetersSZ15,SubercazeGCL16}. Several optimization strategies, e.g., dictionary encoding, balancing workload
and data partitioning, are further studied to enhance parallel RDFS reasoning. There are also parallel implementations for scalable
reasoning of highly expressive ontology languages \cite{SteigmillerLG14,WuH12}. On the other hand, several work utilizes different kinds of computing platforms to make reasoning tasks more efficient in parallel, such as supercomputers \cite{Hoeksema2011,GoodmanJMAAH11},
MapReduce \cite{UrbaniKMHB12} and GPU servers \cite{HeinoP12}.

The above work verifies the effectiveness of parallel reasoning based on the evaluations on different test datasets.
However, for most popular ontology languages, even RDFS and datalog rewritable ontology languages,
which have \texttt{PTime}-complete or higher complexity of reasoning in the worst case,
they are not parallelly tractable according to \cite{Raymond95}, i.e., the efficiency of reasoning may not be
improved on a parallel implementation. A possible reason of the high performance of parallel
reasoning is that the utilized test datasets do not fall in the worst cases in terms of computational complexity.
It has also been shown that some well-known large-scale ontologies, such as YAGO, have good performance for parallel
reasoning \cite{KolovskiWE10}, but they are expressed in ontology languages that are not parallelly tractable in theory.
On the other hand, there do exist ontologies for which materialization cannot
always be improved by parallelism (we use experiments to show this in Section~\ref{sec:evaluation}).
The current work of parallel ontology reasoning can hardly explain this.
While one can try out different parallel implementations to see whether an ontology can be handled by (one of) them efficiently,
we study the problem of parallel tractability in theory and identify properties that make an ontology parallelly tractable.
These properties can be further used to optimize parallel algorithms and to guide ontology engineers in creating ontologies
for which parallel tractability can be guaranteed theoretically.

According to \cite{MotikNPHO14}, many real large-scale ontologies are essentially expressed in the ontology languages
that can be rewritten into datalog rules. Thus, we focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify the classes of datalog rewritable ontologies such that materialization
over these ontologies is parallelly tractable, i.e., in the parallel complexity class NC \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in the NC class, one can give an \emph{NC algorithm} that handles this problem
in parallel computation \cite{Raymond95}.
An NC algorithm is required to terminate in parallel poly-logarithmic time.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not NC algorithms,
since they are designed for general datalog programs and has \texttt{PTime}-complete complexity.
Thus, we first give NC algorithms that perform materialization,
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallelly tractable classes})
that can be handled by these NC algorithms. To make the proposed NC algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite to clarify how to apply the above NC algorithms
to study the parallel tractability of ontology materialization.
We show that DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ are parallelly tractable.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed NC algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze well-known benchmarks and real-world datasets and show that many real ontologies following the
proposed restrictions belong to the parallelly tractable classes.
We implement a system based on an optimized NC algorithm for DHL materialization.
We compare our system to the state-of-the-art reasoner RDFox.
The experimental results show that the optimization proposed in this paper results in a better performance on parallelly
tractable ontologies compared with RDFox.

The rest of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two NC algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study the parallelly tractable materialization of DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto} respectively.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given NC algorithms.
We analyze real-world datasets and evaluate our implementation in Section~\ref{sec:evaluation}.
We then discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}.
The detailed proofs of the theorems and the lemmas in this paper are given in the appendix.

@@

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ can be transformed to a datalog program
(see the corresponding rules in the right column of Table~1).
Note that, for an axiom of the form (T4) $A\sqsubseteq\exists R$, it should be transformed to
a first-order logic rule `$A(x)\rightarrow \exists y(R(x,y))$' where $y$ is called a \emph{free variable} (it is
not occurring in the head atom $A(x)$).
In this work, such a free variable $y$ is eliminated via Skolemisation into a new individual $o_{R}^A$
that corresponds to concept $A$ and role $R$. One can check that the elimination of free variables
does not influence the results of materialization by according to \cite{CalvaneseGLLR07}.

@@

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,...,B_n\rightarrow H\in R, B_i\theta\in\textbf{I} (1\leq i\leq n)\}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.

\subsection{The NC Complexity}

The parallel complexity class NC, known as Nick$'$s
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the NC class
can be solved in poly-logarithmic time on a PRAM (parallel, random-access machine) with 
a polynomial number of processors. We also say that an NC problem can be solved
in \emph{parallel poly-logarithmic time}.
Although the NC complexity is a theoretical analysis tool,
it has been shown that many NC problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, the NC problems are also highly
parallel feasible for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The NC complexity is originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem should be a decision problem in NC.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.

@@

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is parallelly tractable.
Since we assume that for any class of datalog programs where the rule
set of each datalog program is fixed, the materialization problem is thus in data complexity
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}. In other words, the materialization problem on general
datalog programs cannot be solved in parallel poly-logarithmic time unless P=NC.
Thus, we say that \emph{materialization on a class of datalog programs is parallelly tractable
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time (this algorithm is also called an NC algorithm)}.
In this section, we identify such classes of datalog programs by studying different NC algorithms.

@@

We first give the following definition for a parallel tractable class of datalog programs
where an NC algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallelly Tractable Class)} Given a class $\mathcal{D}$ of datalog programs where the
same rule set is shared, we say that $\mathcal{D}$ is a parallelly tractable datalog program (\texttt{PTD}) class
if there exists an NC algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
parallelly tractable ontology (\texttt{PTO}) class.
\end{definition}

According to the above definition, if we find an NC algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not NC algorithms
due to their \texttt{PTime}-complete complexity, since they are designed for handling general datalog programs.
Thus we give our NC algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an NC version
and identify the target \texttt{PTD} class.

@@

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are the two nodes,
we say that node $v_1$ is the parent (node) of node $v_2$ and node $v_2$ is the child (node)
of node $v_1$. Further, $\forall H,B_1,...,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $H$ has a parent node or a child node;
  \item $H$ is the original fact of $P$ if $H$ has the in-degree of $0$;
  \item $B_1,...,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$.
\end{enumerate}
\end{definition}

@@

\subsection{A Basic Parallel Algorithm}
\label{sec:alg-bsc}

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{bsc}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze the parallel
complexity. This model allows us giving \emph{the parallel assumption}: 
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instance in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instances.
Under this assumption, a processor can check the applicability of its corresponding
rule instance and access the state of an atom occurring in this rule instance in constant time.
Algorithm~$\mathsf{A}_{bsc}$ is then given as follows.\\

\noindent\texttt{Algorithm~$\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

@@

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{bsc}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any EDB predicate, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{bsc}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus, 
the whole computing time of Algorithm~$\mathsf{A}_{bsc}$ turns out to be $c_1+l\cdot c_2$
where $l$ denotes the number of iterations of \ref{alg1:updateG}.

When we say that an algorithm is an NC algorithm. It should meet two requirements: first,
it works on a polynomial number of processors; second, it halts in poly-logarithmic time.
As discussed above, Algorithm~$\mathsf{A}_{bsc}$ meets the first requirement.
If we want to restrict Algorithm~$\mathsf{A}_{bsc}$ to be an NC algorithm,
we can make the number of iterations of \ref{alg1:updateG} to be poly-logarithmically bounded.
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
For any datalog program, if we restrict that the number of iterations of \ref{alg1:updateG} is bounded by $\psi(|\textbf{I}|)$, 
the computing time of Algorithm~$\mathsf{A}_{bsc}$
is $c_1+\psi(|\textbf{I}|)\cdot c_2$. With this restriction, Algorithm~$\mathsf{A}_{bsc}$ is an NC algorithm denoted by
$\mathsf{A}_{bsc}^{\psi}$.

@@

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{ex_1}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{ex_1}$.
It can be checked that, for some materialization graph $\mathcal{G}_{ex_1}$
corresponding to a datalog program in $\mathbb{P}_{ex_1}$, \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{ex_1}$ are polynomial of $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{ex_1}$.
However there indeed exists an NC algorithm that can handle $\mathbb{P}_{ex_1}$.
We discuss this in the next part.

>>>>

\address[add2]{
    Institute of Artificial Intelligence, University of Ulm
}

\begin{abstract}
  Materialization is an important reasoning service for many
  ontology-based applications, but the rapid growth of semantic data
  poses the challenge to efficiently perform materialization on
  large-scale ontologies. Parallel materialization algorithms work
  well for some ontologies, which is not always explained by the
  theoretical results. For example, for the YAGO ontology
  parallelization works well, although the reasoning problem for the
  used ontology language is not in \texttt{NC}, i.e.\ the theoretical
  complexity class for \emph{parallel tractability}. This motivates us
  to study the problem of \emph{parallel tractability} of ontology
  materialization from a theoretical perspective. We focus on the
  datalog rewritable ontology languages DL-Lite and Description Horn
  Logic (DHL) and propose algorithms, called \texttt{NC} algorithms,
  to identify classes of ontologies for which materialization is
  tractable in parallel. To verify the practical usability of the
  above results, we analyze real-world datasets and show that for many
  ontologies expressed in DHL materialization is tractable in
  parallel. The implementation of our optimized parallel algorithm
  shows performance improvements over the highly optimized
  state-of-the-art reasoner RDFox on ontologies for which
  materialization is tractable in parallel.
\end{abstract}

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep \texttt{NC} complexity
\end{keyword}

@@

There has been work that proposes parallel reasoning algorithms and employs parallel computing platforms
for the ontology language RDFS and its extended
fragments \cite{MotikNPHO14,PetersSZ15,SubercazeGCL16}. Several optimization strategies, e.g., dictionary encoding, balance workload,
data partitioning, are further studied to enhance parallel RDFS reasoning. There are also parallel implementations for scalable
reasoning of highly expressive ontology languages \cite{SteigmillerLG14,WuH12}. On the other hand, in order to make reasoning tasks more efficient
in parallel, several work utilizes different kinds of high-performance computing platforms, like supercomputers \cite{Hoeksema2011,GoodmanJMAAH11},
MapReduce \cite{UrbaniKMHB12} and GPU servers \cite{HeinoP12}.

The above work verifies the effectiveness of parallel reasoning based on the evaluations on different test datasets.
However, for most popular ontology languages, even RDFS and datalog rewritable ontology languages,
which have \texttt{PTime}-complete or higher complexity of reasoning in the worst case,
they are not parallelly tractable according to \cite{Raymond95}, i.e., the efficiency of reasoning may not be
improved on a parallel implementation. A possible reason of the high performance of parallel
reasoning is that the utilized test datasets do not fall in the worst cases in terms of computational complexity.
It has also been shown that, some well-known large-scale ontologies, such as YAGO, have good performance for parallel
reasoning \cite{KolovskiWE10}, but they are expressed in ontology languages that are not parallelly tractable in theory.
On the other hand, we show in Section~\ref{sec:evaluation} that, there do exist some ontologies, for which materialization cannot
always be improved by parallelism. The current work of parallel ontology reasoning can hardly explain this.
While one can try out different parallel implementations to see whether an ontology can be handled by (one of) them efficiently,
we study the problem of parallel tractability in theory and identify properties that make an ontology parallelly tractable.
These properties can be further used to optimize parallel algorithms and to guide ontology engineers in creating ontologies
for which parallel tractability can be guaranteed theoretically.

According to \cite{MotikNPHO14}, many real large-scale ontologies are essentially expressed in the ontology languages
that can be rewritten into datalog rules. Thus, we focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify the classes of datalog rewritable ontologies such that materialization
over these ontologies is parallelly tractable, i.e., in the parallel complexity class $\texttt{NC}$ \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in the \texttt{NC} class, one can give an \emph{\texttt{NC} algorithm} that handles this problem
in parallel computation \cite{Raymond95}.
An \texttt{NC} algorithm is required to terminate in parallel poly-logarithmic time.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms,
since they are designed for general datalog programs and has \texttt{PTime}-complete complexity.
Thus, we first give \texttt{NC} algorithms that perform materialization,
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallelly tractable classes})
that can be handled by these \texttt{NC} algorithms. To make the proposed \texttt{NC} algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite to clarify how to apply the above \texttt{NC} algorithms
to study the parallel tractability of ontology materialization.
We show that DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ are parallelly tractable.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed \texttt{NC} algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze well-known benchmarks and real-world datasets and show that many real ontologies following the
proposed restrictions belong to the parallelly tractable classes.
We implement a system based on an optimized \texttt{NC} algorithm for DHL materialization.
We compare our system to the state-of-the-art reasoner RDFox.
The experimental results show that the optimization proposed in this paper results in a better performance on parallelly
tractable ontologies compared with RDFox.

The rest of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two \texttt{NC} algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study the parallelly tractable materialization of DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto} respectively.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given \texttt{NC} algorithms.
We analyze real-world datasets and evaluate our implementation in Section~\ref{sec:evaluation}.
We then discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}.
The detailed proofs of the theorems and the lemmas in this paper are given in the appendix.

@@

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ can be transformed to a datalog program
(see the corresponding rules in the right column of Table~1).
Note that, for an axiom of the form (T4) $A\sqsubseteq\exists R$, it should be transformed to
a first-order logic rule `$A(x)\rightarrow \exists y(R(x,y))$' where $y$ is a free variable.
In this work, such a free variable $y$ is eliminated via Skolemisation into a new individual $o_{R}^A$
that corresponds to concept $A$ and role $R$. One can check that the elimination of free variables
does not influence the results of materialization by according to \cite{CalvaneseGLLR07}.

@@

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,...,B_n\rightarrow H\in R, B_i\theta\in\textbf{I} (1\leq i\leq n)\}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}.

\subsection{The $\texttt{NC}$ Complexity}

The parallel complexity class $\texttt{NC}$, known as Nick$'$s
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the $\texttt{NC}$ class
can be solved in poly-logarithmic time on a parallel machine with a polynomial number of processors.
We also say that an $\texttt{NC}$ problem can be solved in \emph{parallel poly-logarithmic time}.
Although the $\texttt{NC}$ complexity is a theoretical analysis tool,
it has been shown that many $\texttt{NC}$ problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, the $\texttt{NC}$ problems are also highly
parallel feasible for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The \texttt{NC} complexity is originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem should be a decision problem in $\texttt{NC}$.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.

@@

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is parallelly tractable.
Since we assume that for any datalog program $\langle R, \textbf{I}\rangle$ the rule
set $R$ is fixed, the materialization problem is thus in data complexity
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}. In other words, the materialization problem on general
datalog programs cannot be solved in parallel poly-logarithmic time unless \texttt{P}=\texttt{NC}.
Thus, we say that \emph{materialization on a class of datalog programs is parallelly tractable
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time (this algorithm is also called an \texttt{NC} algorithm)}.
In this section, we identify such classes of datalog programs by studying different \texttt{NC} algorithms.

@@

We first give the following definition for a parallel tractable class of datalog programs
where an \texttt{NC} algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallelly Tractable Class)} Given a class $\mathcal{D}$ of datalog programs,
we say that $\mathcal{D}$ is a parallelly tractable datalog program (\texttt{PTD}) class
if there exists an \texttt{NC} algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
parallelly tractable ontology (\texttt{PTO}) class.
\end{definition}

According to the above definition, if we find an \texttt{NC} algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms
due to their \texttt{PTime}-complete complexity, since they are designed for handling general datalog programs.
Thus we give our $\texttt{NC}$ algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an \texttt{NC} version
and identify the target \texttt{PTD} class.

@@

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$;
\end{enumerate}
~~$\mathcal{G}$ satisfies the following condition:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $\forall H,B_1,...,B_n\in V$ such that $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$, we have that
  $B_1,...,B_n\rightarrow H\in P^*$.
\end{enumerate}
\end{definition}

@@

\subsection{A Basic Parallel Algorithm}

In this part, we propose a parallel algorithm
(denoted by $\mathsf{A}_{bsc}$) that constructs a materialization graph for a given datalog program.\\

\noindent\texttt{Algorithm $\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Recall that $P^*$ denotes the ground instantiation of $P$,
which consists of all possible ground instantiations of rules in $R$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.\footnote{This might not be practically feasible,
but we focus on a theoretical analysis here.}
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,\footnote{Suppose that each processor
    can use $O(1)$ time units to access the state of ground atoms, i.e., whether this ground
    atom has been added to the materialization graph. This can be implemented by maintaining an
    index of polynomial size.}
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

@@

We now discuss how Algorithm~$\mathsf{A}_{bsc}$ can be restricted to an \texttt{NC} version.
(\uppercase\expandafter{\romannumeral1}) Since Algorithm~$\mathsf{A}_{bsc}$ does not introduce
new constants and each predicate has a constant arity,
one can check that $|P^*|$ is polynomial in the size of $P$. This also means that
the number of processors is polynomially bounded.
(\uppercase\expandafter{\romannumeral2}) The computing time of \ref{alg1:addFacts}
and \ref{alg1:halt} occupies constant time units because of parallelism.
(\uppercase\expandafter{\romannumeral3}) The main computation part in Algorithm~$\mathsf{A}_{bsc}$ is the
iteration of \ref{alg1:updateG}.
In each iteration of \ref{alg1:updateG}, all processors work independently from each other.
Thus, in theory, \ref{alg1:updateG} costs one time unit.
The whole computing time turns out to be bounded by the number of iterations of \ref{alg1:updateG}. (\uppercase\expandafter{\romannumeral4})
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
The input of $\psi$ is the size of $P$ and the output is a non-negative integer.
Based on (\uppercase\expandafter{\romannumeral1}, \uppercase\expandafter{\romannumeral2},
\uppercase\expandafter{\romannumeral3}, \uppercase\expandafter{\romannumeral4}),
for any datalog program $P$, if we use $\psi(|P|)$ to
bound the number of iterations of \ref{alg1:updateG}, then Algorithm~$\mathsf{A}_{bsc}$
is an \texttt{NC} algorithm, denoted by $\mathsf{A}_{bsc}^{\psi}$ (this symbol also
means that $\psi$ occurs as a configuration for Algorithm~$\mathsf{A}_{bsc}$).

@@

The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the
rather simple datalog program $P_{ex_1}$ in Example~\ref{exp:mg}.
The graph $\mathcal{G}_{ex_1}$ in Figure~\ref{fig:mg} is the unique materialization graph of $P_{ex_1}$.
One can also check that \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$.
This means that the depth of $\mathcal{G}_{ex_1}$ is linearly bounded by $k$.
On the other hand, the size of $P_{ex_1}$ is a polynomial of the integer $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph of $P_{ex_1}$.
However there indeed exists an \texttt{NC} algorithm that can handle $P_{ex_1}$.
We discuss this in the next part.
